From DSL to HPC Component Models

### 2015-07-23 WOLFHPC15/article/ms_msc.tex
wolfhpc15 23_07

\begin{figure}[h!]
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
program ::= "mesh:" meshid 
            "mesh domains:" listmeshdom
            "computation domains:" 
                       listcompdom
            "intersections:"
                       listintersection
            "data:" listdata
            "time:" iteration
            "computations:" listcomp
listmeshdom ::= meshdom listmeshdom
             |  meshdom
listcompdom ::= compdom listcompdom
             |  compdom
compdom ::= compdomid "in" listmeshdom
listintersection ::= inter listintersections
                  |  inter
inter ::= compdomid and compdomid
listdata ::= data listdata
          |  data
data ::= dataid "," meshdom
iteration ::= num
listcomp ::= comp listcomp
          |  comp
comp ::= comptype ":" compid "=({" 
            listdataid "}," dataid 
            "," compdomid ")"
listdataid ::= dataid listdataid
            |  dataid
comptype ::= "stencil"
      |  "local"
      |  "bound"
\end{lstlisting}
\caption{Grammar of the MS language}
\label{fig:grammar}
\end{figure}

@@

\paragraph{Computation domains} A computation domain can be seen as a subpart of a meh domain which will be used during computation phase. For each computation domain is indicated the mesh domain identifier from which it is built. After definitions of computation domains, the intersections between those are given. For example
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
computation domains:
  allcell in cell
  alledgex in edgex
  alledgey in edgey
  partedgex in edgex
interactions:
  alledgex and partedgex
\end{lstlisting}
In this example, an entire computation domain is defined for each mesh domain, and additionnally a part of the mesh domain edgex is defined. As a result computation domains alledgex and partedgex intersect.

@@

\paragraph{Computations description} The most important part of MSL is the description of the computations of the simulation. The language follows the definitions given in Section~\ref{sect:multistencil}, however as MSL is not made to express numerical computations, the element $exp$ of the definition is not given in the language. Moreover, an additionnal information is asked to the user, the type of computation. A computation can be \emph{stencil}, \emph{local} or \emph{bound}. \textbf{OU INTRODUIRE bound et refs?}
Thus, the description of a computation consists of a type of computation, an identification name of the computation, a set of identifiers for the data read by the computation, a single identifier for the data written by the computation. Moreover, if the type of computation is a stencil, a neighborhood identifier is needed.
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
computations:
  local:c0({a},b)
  stencil:c1({b},c,n1)
  local:c2({c},d)
  local:c3({c},e)
  stencil:c4({d},f)
  local:c5({e},g)
  local:c6({f},h)
  local:c7({g,h},i)
  stencil:c8({i},j,n1)
\end{lstlisting}

@@

\paragraph{Creation of the ordered list of computations} This step of the MSC compiler parse the MSL input file which describes the simulation. From the ordered lines of computations, it builds the ordered list of computations $\Gamma$ to perform in the simulation.

\paragraph{Creation of the synchronized ordered list of computations} From the ordered list of computations $\Gamma$ it is possible to detect needed synchronizations if a \emph{data parallelization} is performed on the simulation. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation different from a boundary computation. Actually, as a stencil computation needs neighborhood values, in a domain decomposition values computed by different processors or cores are needed.

\paragraph{Creation of the dependency graph} From the synchronized ordered list of computations $\Gamma_{data}$, a dependency graph can be build. A dependency exists between two computations (including synchronizations) if and only if a data read has been written by a previous computation in $\Gamma_{data}$. Nodes of the dependency graph are computations and synchronizations, while edges are dependencies between them. $\Gamma_{hybrid}$ is a directed acyclic graph (\emph{dag}) as it is build from an ordered list.

\paragraph{Transformation to a minimal series-parallel graph} As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a dag is a minimal series-parallel graph if and only if the forbidden \emph{N-shape} is not found in the graph. To transform $\Gamma_{hybrid}$ to the minimal series-parallel graph $\Gamma_{msp}$, a BFS algorithm is applied to build a complete bipartite between two levels of $\Gamma_{hybrid}$~\cite{Mitchell:2004:CMV:1082101.1082117}.

\paragraph{Creation of a tree decomposition} As shown in different works~\cite{Valdes:1979:RSP:800135.804393,Schoenmakers95anew}, from a minimal series-parallel graph a tree decomposition can be built. A series-parallel tree decomposition consists in the decomposition of the graph as a set of \emph{sequences} and \emph{parallel} sections in a tree.

For more details on MSC, a full description of the compilation is described in~\cite{}.

%----------------------------------------
\subsection{Example}
%----------------------------------------
In this Section we give an example of an input file written with MSL, the $\Gamma_{tsp}$ series-parallel tree decomposition computed and the final component dump.

The input MSL file, shown in Figure~\ref{fig:msinput} is composed of nine computations and ten data.

\begin{figure}[h!]
\begin{lstlisting}[basicstyle=\footnotesize,mathescape,frame=single,language=C++]
data:
  a,d1
  b,d1
  c,d2
  d,d2
  e,d3
  f,d1
  g,d3
  h,d2
  i,d1
  j,d2
time:500
computations:
  local:c0({a},b)
  stencil:c1({b},c,n1)
  local:c2({c},d)
  local:c3({c},e)
  stencil:c4({d},f)
  local:c5({e},g)
  local:c6({f},h)
  local:c7({g,h},i)
  stencil:c8({i},j,n1)
\end{lstlisting}
\caption{Example of MSL input file}
\label{fig:msinput}
\end{figure}

The first step of the compilation is to build the synchronized ordered list of computations $\Gamma_{data}$. In this example each stencil computation actually needs a synchronization. For example, the stencil computation $c_1$ read the data $b$ which has been written by the computation $c_0$. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed to the sublist $[c_0,sync_1,c_1]$ in $\Gamma_{data}$. The new computation $sync_1$ read and write $b$. As a result a dependency is kept between $sync_1$, and $c_1$. The same is performed for the stencils $c_4$ and $c_8$.

Once the synchronized list $\Gamma_{data}$ is obtained, the dependency graph $\Gamma_{hybrid}$ is built and illustrated in Figure~\ref{fig:hyb}. Each computation (and synchronization) is a node of the graph, while each edge represent a read/write data dependency between computations.

@@

After a set of transformation briefly explained in the previous Section, a series-parallel tree decomposition $\Gamma_{tsp}$ is obtained and illustrated in Figure~\ref{fig:canon}.

\begin{figure}[h!]
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[] (s0) at (0,0) {$\mathcal{S}$};
   \node[] (c0) at (-3,1) {$c_0$};
   \node[] (star1) at (-2,1) {$sync_1$};
   \node[] (c1) at (-1,1) {$c_1^*$};

   \node[] (p0) at (0,1) {$\mathcal{P}$};
   \node[] (s1) at (-1,2) {$\mathcal{S}$};
   \node[] (p1) at (-2,3) {$\mathcal{P}$};
   \node[] (c4) at (-1,3) {$c_4^*$};
   \node[] (c6) at (-0,3) {$c_6$};
   \node[] (c2) at (-2.5,4) {$c_2$};
   \node[] (star4) at (-1.5,4) {$sync_4$};
   \node[] (s2) at (1,2) {$\mathcal{S}$};
   \node[] (c3) at (0.5,3) {$c_3$};
   \node[] (c5) at (1.5,3) {$c_5$};

   \node[] (c7) at (1,1) {$c_7$};
   \node[] (star8) at (2,1) {$sync_8$};
   \node[] (c8) at (3,1) {$c_8^*$};

@@

>>>>

\begin{figure}[h!]
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
program ::= "mesh:" meshid 
            "mesh domains:" listmeshdom
            "computation domains:" 
                       listcompdom
            "no intersections:"
                       listnointer
            "data:" listdata
            "time:" iteration
            "computations:" listcomp
listmeshdom ::= meshdom listmeshdom
             |  meshdom
listcompdom ::= compdom listcompdom
             |  compdom
compdom ::= compdomid "in" listmeshdom
listnointer ::= nointer listnointer
                  |  nointer
nointer ::= compdomid and compdomid
listdata ::= data listdata
          |  data
data ::= dataid "," meshdom
iteration ::= num
listcomp ::= comp listcomp
          |  comp
comp ::= dataid "=" compid "=({" 
            listdataid "}," compdomid 
            ["," neighborid] ")"
listdataid ::= dataid listdataid
            |  dataid
comptype ::= "stencil"
      |  "local"
      |  "bound"
\end{lstlisting}
\caption{Grammar of the MS language}
\label{fig:grammar}
\end{figure}

@@

\paragraph{Computation domains} A computation domain can be seen as a subpart of a meh domain which will be used during computation phase. For each computation domain is indicated the mesh domain identifier from which it is built. By default two computation domains always intersect except when precise in the section \emph{no intersections}. For example
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
computation domains:
  allcell in cell
  alledgex in edgex
  alledgey in edgey
  part1edgex in edgex
  part2edgex in edgex
no intersections:
  part1edgex and part2edgex
\end{lstlisting}
In this example, an entire computation domain is defined for each mesh domain, and additionnally two subpart of the mesh domain edgex are defined. It is then precised that part1edgex and part2edgex do not intersect.

@@

\paragraph{Computations description} The most important part of MSL is the description of the computations of the simulation. The language follows the definitions given in Section~\ref{sect:multistencil}, however as MSL is not made to express numerical computations, the element $exp$ of the definition is not given in the language. Moreover, an additionnal information is asked to the user, the type of computation. A computation can be \emph{stencil}, \emph{local} and if a stencil is computed a neighborhood identifier is given.
Thus, the description of a computation consists of a type of computation, an identification name of the computation, a set of identifiers for the data read by the computation, a single identifier for the data written by the computation. Moreover, if the type of computation is a stencil, a neighborhood identifier is needed.
\begin{lstlisting}[basicstyle=\small,mathescape,frame=single,language=C++]
computations:
  b=c0({a},allcell)
  c=c1({b},alledgex,n1)
  d=c2({c},alledgex)
  e=c3({c},alledgey)
  f=c4({d},allcell,n1)
  g=c5({e},alledgey)
  h=c6({f},alledgex)
  i=c7({g,h},allcell)
  j=c8({i},partedgex,n1)
\end{lstlisting}

@@

\paragraph{Creation of the ordered list of computations} This step of the MSC compiler parse the MSL input file which describes the simulation. From the ordered lines of computations, it builds the ordered list of computations $\Gamma$ to perform in the simulation. For example the ordered list of computation of the example above is $[c_0,c_1,c_2,c_3,c_4,c_5,c_6,c_7,c_8]$.

\paragraph{Creation of the synchronized ordered list of computations} From the ordered list of computations $\Gamma$ it is possible to detect needed synchronizations if a \emph{data parallelization} is performed on the simulation. A synchronization is needed each time a data read by a stencil computation has been written by a previous computation and if the computation domains intersect. Actually, as a stencil computation needs neighborhood values, in a domain decomposition values computed by different processors or cores are needed.

For example, the stencil computation $c_1$ read the data $b$ which has been written by the computation $c_0$. Moreover the mesh domains are different which means that the computation domains intersect. For this reason the sublist $[c_0,c_1]$ of $\Gamma$ is transformed to the sublist $[c_0,sync_1,c_1]$ in $\Gamma_{data}$. The new computation $sync_1$ read and write $b$. As a result a dependency is kept between $sync_1$, and $c_1$. The same is performed for the stencils $c_4$ and $c_8$.

\paragraph{Creation of the dependency graph} From the synchronized ordered list of computations $\Gamma_{data}$, a dependency graph can be build. A dependency exists between two computations (including synchronizations) if and only if a data read has been written by a previous computation in $\Gamma_{data}$, and if the computation domains intersect. In other words, the only case where a data dependency exists but is not taking into account is when the computation domains do not intersect. Nodes of the dependency graph represent computations and synchronizations, while edges are dependencies between them. $\Gamma_{hybrid}$ is a directed acyclic graph (\emph{dag}) as it is build from an ordered list. For example, the dependency dag $\Gamma_{hybrid}$ of the example is illustrated in Figure~\ref{fig:hyb}.

@@

\paragraph{Transformation to a minimal series-parallel graph} As it has been shown in~\cite{Valdes:1979:RSP:800135.804393}, the transitive reduction of a dag is a minimal series-parallel graph if and only if the forbidden \emph{N-shape} is not found in the graph. To transform $\Gamma_{hybrid}$ to the minimal series-parallel graph $\Gamma_{msp}$, a breadth first search~\cite{} traversing algorithm is applied to build a complete bipartite graph between two levels of $\Gamma_{hybrid}$~\cite{Mitchell:2004:CMV:1082101.1082117}.

\paragraph{Creation of a tree decomposition} As shown in different works~\cite{Valdes:1979:RSP:800135.804393,Schoenmakers95anew}, from a minimal series-parallel graph a tree decomposition can be built. A series-parallel tree decomposition consists in the decomposition of the graph as a set of \emph{sequences} and \emph{parallel} sections in a tree. For example, the series-parallel tree decomposition $\Gamma_{tsp}$ of the example is illustrated in Figure~\ref{fig:canon}.

\begin{figure}[h!]
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[] (s0) at (0,0) {$\mathcal{S}$};
   \node[] (c0) at (-3,1) {$c_0$};
   \node[] (star1) at (-2,1) {$sync_1$};
   \node[] (c1) at (-1,1) {$c_1$};

   \node[] (p0) at (0,1) {$\mathcal{P}$};
   \node[] (s1) at (-1,2) {$\mathcal{S}$};
   \node[] (p1) at (-2,3) {$\mathcal{P}$};
   \node[] (c4) at (-1,3) {$c_4$};
   \node[] (c6) at (-0,3) {$c_6$};
   \node[] (c2) at (-2.5,4) {$c_2$};
   \node[] (star4) at (-1.5,4) {$sync_4$};
   \node[] (s2) at (1,2) {$\mathcal{S}$};
   \node[] (c3) at (0.5,3) {$c_3$};
   \node[] (c5) at (1.5,3) {$c_5$};

   \node[] (c7) at (1,1) {$c_7$};
   \node[] (star8) at (2,1) {$sync_8$};
   \node[] (c8) at (3,1) {$c_8$};

@@

For more details on MSC, a full description of the compilation is described in the research report~\cite{}.
